### 2013-05-06 article.tex
Final proof read
Signed-off-by: Michael Barton <mail@michaelbarton.me.uk>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. Yet there remains a steep
learning curve in the subsequent computational steps required to process raw
sequence reads into a database-deposited genome sequence. Here we describe
``Genomer,'' a tool to simplify the manual tasks of finishing and uploading a
genome sequence to a database. Genomer can format a genome scaffold into the
common files required for submission to GenBank. This software also simplifies
updating a genome scaffold by allowing a human-readable YAML format file to be
edited instead of large sequence files. Genomer is written as a command line
tool and is an effort to make the manual process of genome scaffolding more
robust and reproducible. Extensive documentation and video tutorials are
available at http://next.gs.

@@

The decreasing costs and increasing diversity of high-throughput sequencing
methods is making genome analysis a common method to tackle unresolved
questions in microbiology \cite{loman2012b}. Sequencing can produce
$10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
larger contigs which are then scaffolded into larger megabase fragments or
complete chromosomes. These larger DNA sequences are then annotated with
genomic features such as protein-coding genes. Each of these required steps can
however produce imperfect results and remains the subject of active research
\cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
sequence and corresponding annotations may still improve upon the results of
automated methods.

@@

We anticipate these problems have resulted in individual groups independently
producing in-house programming scripts to automate repetitive tasks. An example
of such tasks include finding protein names that do not match a convention of
three lower case characters followed by one upper case character (abcD). Such
tasks are manually laborious but simple to automate computationally. Individual
research groups producing their own scripts for such tasks however leads to a
repetition of effort. In contrast open-source projects can lead to a pooling of
community effort resulting in a higher standard of software and preventing
reinvention of the same code \cite{ince2012}.

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer,'' to ease and simplify many of the
common steps in our genome project. While several graphical user interface
tools are available to work with genome projects, including G-InforBio,
Genquire, Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002,
lopez2011, carver2012, gordon1998}. Genomer provides a simple command line
interface which allows scriptable automation and reproducible results. We made
this software open-source under the MIT license and here we describe its
application.

@@

  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling draft genome
  sequences. This is format requires only the order and IDs of each contig be
  specified. This thereby simplifies the process of re-organising and trimming
  contigs in draft genome without having to edit large nucleotide sequences
  manually.}

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to their respective positions in the draft
  genome sequence. This allows the scaffold file to be continuously improved
  and updated even after the contigs have been annotated. For instance,
  additional PCR sequences can be used to close gaps in the assembly and the
  location of all annotations be correspondingly updated.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the scaffold file and corresponding
  GFF3 file and removes the effort to produce these manually.}

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Install
  plugins may also be locked to specific versions to prevent backwards
  incompatible software changes breaking an existing workflow.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool around plain text files. This allows for integration with common Linux
  tools tools such as GNU Make or git thereby allowing reproducible scripting
  of the genome finishing process.}

@@

We have released Genomer as open source software on GitHub
\cite{genomer-github} and have prepared documentation and video tutorials at
http://next.gs. We have used this tool extensively to simplify our own genome
projects and believe Genomer may also be useful to others in the field. We will
illustrate the possible application of Genomer with the following example use
case.

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs and
  unresolved gap regions in the scaffold. Paired-read sequencing or comparison
  to reference genomes can be used as a source of contig orientation and order.
  As an example, in our own genome projects we determine contig order by
  aligning contigs to a reference genome using nucmer \cite{kurtz2004}. A
  detailed overview of the scaffolder file format can be found on the Genomer
  website.}

@@

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and \emph{in silico} analysis. Genomer allows
  changes in the scaffold file to automatically propagate to the required
  changes in the generated files when the \verb+genomer view+ command is
  rerun.}

  \item{The required files for submission to GenBank may be generated when
  researchers are satisfied with the status of their draft. These files include
  the assembled FASTA file, contig FASTA file, annotation table file, and AGP
  scaffold file. Additional options can be used to add the required prefix to
  gene annotation IDs and begin locus ID numbering at the origin of the
  sequence. Genomer does not produce the .asn file which is submitted to
  GenBank, but instead the files that may be used as the input to tbl2asn or
  sequin provided by GenBank.}

@@

Genomer is an open-source tool to simplify and automate repetitive and
time-consuming tasks required when finishing a microbial genome project. This
software is available on GitHub \cite{genomer-github} with documentation and
video tutorials on the http://next.gs website. This tool has been useful in our
own genome projects and we believe it will also be useful to other researchers
especially smaller research groups entering the field for the first time.

>>>>

The increasing accessibility and reduced costs of sequencing has made genome
analysis accessible to more and more researchers. Yet there remains a steep
learning curve in the subsequent computational steps required to process raw
reads into a database-deposited genome sequence. Here we describe ``Genomer,''
a tool to simplify the manual tasks of finishing and uploading a genome
sequence to a database. Genomer can format a genome scaffold into the common
files required for submission to GenBank. This software also simplifies
updating a genome scaffold by allowing a human-readable YAML format file to be
edited instead of large sequence files. Genomer is written as a command line
tool and is an effort to make the manual process of genome scaffolding more
robust and reproducible. Extensive documentation and video tutorials are
available at http://next.gs.

@@

The decreasing costs and increasing diversity of high-throughput sequencing
methods is making genome analysis a common method to tackle unresolved
questions in microbiology \cite{loman2012b}. Sequencing can produce
$10e^{6}$-$10e^{10}$ short nucleotide reads, which must be assembled into
larger contigs that are scaffolded into larger megabase fragments or complete
chromosomes. These larger DNA sequences are then annotated with genomic
features such as protein-coding genes. However each of these required steps can
produce imperfect results and remains the subject of active research
\cite{earl2011,quail2012,beckloff2012}. Therefore manual curation of a genome
sequence and corresponding annotations may still improve upon the results of
automated methods.

@@

We anticipate these problems have resulted in individual groups independently
producing in-house programming scripts to automate repetitive tasks. An example
of such tasks include finding protein names that do not match a convention of
three lower case characters followed by one upper case character (abcD). Such
tasks are manually laborious but simple to automate computationally. Individual
research groups producing their own scripts for such tasks however leads to a
repetition of effort. In contrast open-source projects can lead to a pooling of
community effort resulting in a higher standard of software and prevents
reinvention of the same code \cite{ince2012}.

During the process of sequencing several \emph{Pseudomonas fluorescens} strains
we developed a tool, called ``Genomer,'' to ease and simplify many of the
common steps in our genome project. Several graphical user interface tools are
available to work with genome projects, including G-InforBio, Genquire,
Pile-LineGUI, Artemis and Consed \cite{tanaka2006, wilkinson2002, lopez2011,
carver2012, gordon1998}. Genomer instead provides a simple command line
interface which allows scriptable automation and reproducible results. We made
this software open-source under the MIT license and here we describe its
application.

@@

  \item[Simple editing of a draft genome sequence.]{Genomer is built on the
  existing Scaffolder \cite{barton2012} file format for assembling draft genome
  sequences. This is format requires only the order and IDs of each contig be
  specified. This thereby simplifies the process of re-organising and trimming
  contigs in a draft genome without having to edit large nucleotide sequences
  manually.}

  \item[Mapping of annotations onto the assembled sequence.]{Genomer maps the
  coordinates of contig annotations to their respective positions in the draft
  genome sequence. This allows the scaffold file to be continuously improved
  and updated even after the contigs have been annotated. For instance,
  additional PCR sequences can be used to close gaps in the assembly and the
  location of all annotations will be correspondingly updated.}

  \item[Generation of files for submission to GenBank.]{Submitting a genome
  sequence to GenBank database requires generating specific files. These may
  include a FASTA file of the draft sequence, a table of annotations, a FASTA
  file containing the individual contig sequences, and `a golden path' (AGP)
  \cite{agp-spec} file describing the placement of contigs. Genomer automates
  the generation of all these files from the scaffold file and corresponding
  GFF3 file thereby removing the effort to produce these manually.}

  \item[A stable interface and streamlined install process]{Genomer provides a
  plugin system built using the existing Ruby package management system:
  RubyGems and Bundler. This allows Genomer and its plugins to be automatically
  installed without requiring manual downloading and compilation by the user.
  This eliminates a common problem in bioinformatics where software may require
  technical expertise to first compile and install before being used. Installed
  plugins may also be locked to specific versions to prevent backwards
  incompatible software changes breaking an existing workflow.}

  \item[Integration with the command line.]{Genomer is built as a command line
  tool around plain text files. This allows for integration with common Linux
  tools tools such as GNU Make or git allowing reproducible scripting of the
  genome finishing process.}

@@

We have released Genomer as open source software on GitHub
\cite{genomer-github} and have prepared documentation and video tutorials at
http://next.gs. We have used this tool extensively to simplify our own genome
projects and believe Genomer may also be useful to others in the field. We will
illustrate the possible application of Genomer with the following use case.

@@

  \item{A genome scaffold is written in the scaffolder file format
  \cite{barton2012}. This specifies the order and orientation of contigs and
  unresolved gap regions in the scaffold. Paired-read sequencing or comparison
  to reference genomes can be used as a source of contig orientation and order.
  As an example, in our own genome projects we determine contig order by
  aligning to a reference genome using nucmer \cite{kurtz2004}. A detailed
  overview of the scaffolder file format can be found on the Genomer website.}

@@

  \item{The genome scaffold can be continuously updated through closing any
  remaining gaps. During our microbe sequencing projects we closed gaps through
  a combination of Sanger PCR and \emph{in silico} analysis. Genomer allows
  changes in the scaffold file to automatically propagate to the generated
  files when the \verb+genomer view+ command is rerun.}

  \item{The required files for submission to GenBank may be generated when
  researchers are satisfied with the status of their draft. These files include
  the assembled FASTA file, contig FASTA file, annotation table file, and AGP
  scaffold file. Additional options can be used to add the required prefix to
  gene annotation IDs and begin locus ID numbering at the origin of the
  sequence. Genomer does not produce the .asn file which is submitted to
  GenBank, but instead the files that are used as the input to tbl2asn or
  sequin tools provided by GenBank.}

@@

Genomer is an open-source tool to simplify and automate repetitive and
time-consuming tasks required when finishing a microbial genome project. This
software is available on GitHub \cite{genomer-github} with documentation and
video tutorials on the http://next.gs website. This tool has been useful in our
own genome projects, and we believe it will also be useful to other researchers
especially smaller research groups entering the field for the first time.
